---
title: "rrapidmarkov: User Guide"
author: "Ed Wilson"
date: "`r Sys.Date()`"

output: 
  rmarkdown::html_vignette: 
    toc: true
  
vignette: >
  %\VignetteIndexEntry{User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
This package is designed as a 'no frills' Markov model calculator primarily
aimed at health economists conducting cost-effectiveness or cost-utility 
analyses for health technology assessment.  Whilst there are some great 
comprehensive packages on [CRAN](https://cran.r-project.org/) for 
decision modelling (eg [heemod](https://CRAN.R-project.org/package=heemod)), I
wanted a basic, stripped down calculator that could rapidly compute Markov 
models receiving input and generating output in my own preferred, somewhat 
idiosyncratic, format.

The focus of this package is on speed of computation, and the results of 
benchmarking tests as well as due-diligence accuracy checks of the code are in 
the [accompanying]("validation-and-speed-check.html") vignette.

Output can easily be imported into packages such as [BCEA](https://CRAN.R-project.org/package=BCEA) and [SAVI](https://github.com/Sheffield-Accelerated-VoI/SAVI) to take advantage
of their functionality, such as Value of Information analysis calculations.

I hope you find this useful in your own analyses.
<hr>

# Getting started

Before looking at some examples, a few points to note:

## Terminology

The **state-transitions matrix** is a square matrix (that is, it has
an equal number of rows and columns), showing the probability of transitioning
from one state to another.  The number of rows and columns is equal
to the number of health states.  The rows represent the 'from' state and the
columns the 'to state', thus cell [1,2] of the matrix is the probability of
going from state 1 to state 2 (0.1 in the example below).  The rows of the 
matrix must sum to 1.

```{r echo = FALSE}
cat("Example state-transitions matrix:\n")
stateN <- c("State 1", "State 2", "State 3")
matrix(data = c(0.8, 0.1, 0.1,
                0.0, 0.7, 0.3,
                0.0, 0.0, 1.0), 
       ncol = 3, byrow = TRUE,
       dimnames = list(stateN, stateN))
```

A **stationary** Markov model (sometimes called **temporally stationary**)
is where the transition probabilities do not vary with time.  A 
**non-stationary** Markov model is where the transition probabilities are 
different each cycle, for example to reflect increasing probability of death 
with age, or changing hazard of disease recurrence in the years after definitive
surgery etc.

It follows that defining a stationary model needs only a single 
state-transition matrix to be specified, but for a non-stationary model, a 
separate matrix needs defining for each cycle (or rather, a 3d array where the 
'slices' (z-dimension) represent cycle number.  The examples below show how to 
do this using either life tables or a survival curve as the source data.

## Cycles are numbered from zero

The state of the Markov chain in t=0 is defined exogenously, so for a 
non-stationary model (i.e. where transition probabilities vary over time), 
you must define (cycles-1) transition matrices.

## Markov models should always be analysed probabilistically

As Markov models are non-linear, running a single iteration with the function
runmarkov(), with all input parameters at their means will not yield the
expected state of the model.  The means from a probabilistic analysis
(from psamarkov()) are, by definition the expected states, and 
should be presented as the 'base case' in any economic evaluation.

The functions psamarkov() and psavalues() calculate the means based on repeated
calls to runmarkov() and calcvalues() respectively.  However, during model 
development it is very useful to look at deterministic results for debugging 
and error checking, and so the deterministic functions runmarkov() and 
calcvalues() are exposed / exported for this purpose.  The examples start with
such a deterministic analysis, before moving on to probabilistic analyses.

<hr>

# Examples
After installing the package, add it to your script with the following command:
```{r setup}
library(rrapidmarkov)
```

## Example 1a: My First Markov Model

This example firstly calculates the Markov trace (proportions of cohort in each
state each cycle) then applies costs and QALYs for a three state, temporally
stationary model over ten years with cycle length = 1 year.

The states represent a typical 3-state model, and are labelled 'alive', 
'progressive' and 'dead'.

'startingStates' must be a vector of length n where n = number of states and 
sum(startingStates) = 1.

First the Markov trace:

```{r}
states <- c("alive", "progressive", "dead")
startingStates <- c(1, 0, 0)
cycles <- 10
transitionsMatrix <- matrix(data = c(0.8, 0.1, 0.1,
                                     0.0, 0.7, 0.3,
                                     0.0, 0.0, 1.0), ncol = 3, byrow = TRUE)

output <- runmarkov(transitionsMatrix, startingStates, cycles, states)
print(output)
```

Next apply the costs and health state utilities.  Assume the annnual cost of 
'alive' is £100, 'progressive' is £1000 and 'dead' is zero.  Health state 
utilities are 0.8, 0.5 and 0.0 respectively. The transition period (cycle 
length) is 1 year, and the discount rate is 3.5% pa.

```{r}
# apply costs and outcomes
stateCosts <- c(100, 1000, 0)
stateUtilities <- c(0.8, 0.5, 0)
transPeriod <- 1
discountRate <- 0.035
```
When calculating QALYs from health state utilities, set the "QALYs" flag to 
TRUE.  This ensures the calculations are adjusted for the transition period
(see Example 1b).
```{r results="hide"}
costs <- calcvalues(output, stateCosts, transPeriod, discountRate)
QALYs <- calcvalues(output, stateUtilities, transPeriod, discountRate, 
                    QALYs = TRUE)
```
```{r}
print(costs)
print(QALYs)
```

## Example 1b: Six-monthly transition period

This example is identical to 1a, but with a six-monthly rather than annual 
transition period.  

transPeriod should be entered as a fraction of a year.  Eg day = 1/365, week = 
1/52, calendar month = 1/12, lunar (4-week) month = 1/13, quarter = 1/4 and so
on.  This is best entered as a fraction rather than decimal.
```{r results="hide"}
states <- c("alive", "progressive", "dead")
startingStates <- c(1, 0, 0)
cycles <- 10
transitionsMatrix <- matrix(data = c(0.8, 0.1, 0.1,
                                     0.0, 0.7, 0.3,
                                     0.0, 0.0, 1.0), ncol = 3, byrow = TRUE)
output <- runmarkov(transitionsMatrix, startingStates, cycles, states)
stateCosts <- c(100, 1000, 0)
stateUtilities <- c(0.8, 0.5, 0)
transPeriod <- 1/2
discountRate <- 0.035
costs <- calcvalues(output, stateCosts, transPeriod, discountRate)
QALYs <- calcvalues(output, stateUtilities, transPeriod, discountRate, 
                    QALYs = TRUE)
```
Compare the results below with Example 1a.  Undiscounted costs are unchanged but 
undiscounted QALYs are halved.  Discounted costs and QALYs take into account the 
shorter transition period.

By convention first year costs and outcomes are not discounted.  For transition
periods of less than one year (i.e. where transPeriod<1), this means the 
discount factor for the first 1/transPeriod cycles will be 1.  Subsequent cycles
are discounted by 1/((1+r)^(transPeriod x t)), so for a discountRate of 3.5% and
a transPeriod of 1/2 (i.e. 6 months), cycles 0 and 1 will not be discounted,
cycle 2 has a discount factor of 1/((1+0.035)^(0.5x2)) = 0.966, cycle 3 has a
discount factor of 1/((1+0.035)^(0.5x3)) = 0.950 and so on.

Note the code also works where the transition period is greater than one year,
with QALYs and discounting adjusted accordingly.

```{r}
print(costs)
print(QALYs)
```

## Example 2a: A non-stationary model

This example calculates the cost and QALYs from a three-state non-stationary
model (probabilities vary with time) over six cycles (cycle length = 1 year).

Note the 'dim' argument of array() for defining transitionsMatrix must equal
(nStates, nStates, cycles-1), so here it is c(3, 3, 5).

aperm() transposes the array to the correct structure.

```{r}
states <- c("alive", "progressive", "dead")
startingStates <- c(1, 0, 0)
transitionsMatrix <- aperm(array(data = c(0.8, 0.1, 0.1,
                                          0.0, 0.7, 0.3,
                                          0.0, 0.0, 1.0,

                                          0.7, 0.1, 0.2,
                                          0.0, 0.6, 0.4,
                                          0.0, 0.0, 1.0,

                                          0.6, 0.1, 0.3,
                                          0.0, 0.5, 0.5,
                                          0.0, 0.0, 1.0,

                                          0.5, 0.1, 0.4,
                                          0.0, 0.4, 0.6,
                                          0.0, 0.0, 1.0,

                                          0.4, 0.1, 0.5,
                                          0.0, 0.3, 0.7,
                                          0.0, 0.0, 1.0),
                                 dim = c(3, 3, 5)), 
                           c(2, 1, 3))
output <- runmarkov(transitionsMatrix, startingStates, stateNames = states)
print(output)
```
Costs and QALYs can be applied by running the 'calcvalues' function as described
in Example 1a:

```{r results="hide"}
stateCosts <- c(100, 1000, 0)
stateUtilities <- c(0.8, 0.5, 0)
costs <- calcvalues(output, stateCosts)
QALYs <- calcvalues(output, stateUtilities, QALYs = TRUE)
```
```{r}
print(costs)
print(QALYs)
```

## Example 2b: Another non-stationary model

Two state, non-stationary model, state names not specified, 50/50 
split between states at t=0, run for four cycles.

```{r}
startingStates <- c(0.5, 0.5)
transitionsMatrix <- aperm(array(data = c(0.8, 0.2,
                                          0.3, 0.7,

                                          1, 0,
                                          0, 1,

                                          0, 1,
                                          0.5, 0.5),
                                 dim = c(2, 2, 3)), 
                           c(2, 1, 3))

output <- runmarkov(transitionsMatrix, startingStates)
print(output)
```

Again, costs and QALYs can be applied by running the 'calcvalues' function as
described in Example 1a:
```{r results = "hide"}
stateCosts <- c(100, 1000)
stateUtilities <- c(0.8, 0)
costs <- calcvalues(output, stateCosts)
QALYs <- calcvalues(output, stateUtilities, QALYs = TRUE)
```
```{r}
print(costs)
print(QALYs)
```

## Example 3: Using the converttomatrices() helper function

Directly defining the transitionsMatrix is a little unwieldy for probabilistic 
analyses (Monte Carlo simulation), especially for non-stationary models. My 
preferred format when sampling is for every row of a table (which I call 
'probs') to contain a complete set of all probabilities used in one iteration 
of a model.  In a non-stationary Markov model, this leads to a lot of columns 
(nStates^2 x (cycles-1)), but I find the neatness of one row per simulation 
outweighs this.  However, psamarkov() requires a list with a 3d transitions 
matrix for each simulation as a separate item.

converttomatrices() converts the flat 'probs' table to a list where each list 
item contains the state-transitions matrix for each Monte Carlo simulation. This
can be used as the transitionsMatrix parameter passed to psamarkov().

### Example 3a: Stationary model
probs must contain a set of sampled transition probabilities as a matrix of 
size n x c, where n = number of Monte Carlo simulations and c = nStates^2 
(where nStates = number of health states).

The column order for, say, a 2 state model must be: FromState1toState1, 
FromState1toState2, FromState2toState1, FromState2toState2.

This example is a model with three states, run for three cycles. Transition
probabilities are sampled from Dirichlet distributions using 
gtools::rdirichlet().

```{r results = "hide"}
library(gtools)
sims <- 10
set.seed(34)
probs <- cbind(rdirichlet(sims,c(80, 10, 10)),
               rdirichlet(sims,c(0, 7, 3)),
               rdirichlet(sims,c(0, 0, 1)))
colnames(probs) <- c("S1>S1", "S1>S2", "S1>S3",
                     "S2>S1", "S2>S2", "S2>S3",
                     "S3>S1", "S3>S2", "S3>S3")
states <- c("alive", "progressive", "dead")
cycles <- 3
transitionsMatrix <- converttomatrices(states, probs, cycles)
```
There are *sims* sets of matrices (here sims=10).  The first two are shown below.
Note that within each *sim*, the transitions matrices for each cycle are
identical: this is a stationary model, but the matrices are different between
each sim, representing a possible realisation of the world based on the 
Dirichlet samples.
```{r}
transitionsMatrix[1:2]
```

### Example 3b: Non-stationary model
probs must contain a set of sampled transition probabilities as a matrix of 
size n x c, where n = number of Monte Carlo simulations and c = 
(nStates^2)*(cycles-1) columns, (where nStates = number of health states, cycles
= number of cycles).

The column order for, say, a 2 state model with 3 cycles must be:
Cycle1FromState1toState1, Cycle1FromState1toState2,
Cycle1FromState2toState1, Cycle1FromState2toState2,
Cycle2FromState1toState1, Cycle2FromState1toState2,
Cycle2FromState2toState1, Cycle2FromState2toState2.

This example is a model with three states, run for three cycles. Transition
probabilities are sampled from Dirichlet distributions using 
gtools::rdirichlet(), with a different distribution specified for each cycle.
Specifically, there is a 10% probability of moving from 'alive' to 'progressive'
in cycle 1, rising to 20% in cycle 2.

```{r results = "hide"}
library(gtools)
sims <- 10
set.seed(34)
probs = cbind(rdirichlet(sims,c(80, 10, 10)),
             rdirichlet(sims,c(0, 7, 3)),
             rdirichlet(sims,c(0, 0, 1)),

             rdirichlet(sims,c(70, 20, 10)),
             rdirichlet(sims,c(0, 7, 3)),
             rdirichlet(sims,c(0, 0, 1)))

colnames(probs) <- c("c1_S1>S1", "c1_S1>S2", "c1_S1>S3",
                     "c1_S2>S1", "c1_S2>S2", "c1_S2>S3",
                     "c1_S3>S1", "c1_S3>S2", "c1_S3>S3",

                    "c2_S1>S1", "c2_S1>S2", "c2_S1>S3",
                    "c2_S2>S1", "c2_S2>S2", "c2_S2>S3",
                    "c2_S3>S1", "c2_S3>S2", "c2_S3>S3")
states <- c("alive", "progressive", "dead")
transitionsMatrix <- converttomatrices(states, probs)
```
transitionsMatrix contains *sims* sets of matrices (here sims=10).  The first 
two are shown below.  Note that within each *sim*, the transitions matrices for 
each cycle are different: this is a non-stationary model.  The sets of matrices
are different between each sim, representing a possible realisation of the world 
based on the input Dirichlet distributions.

```{r}
transitionsMatrix[1:2]
```

## Example 4: Bringing everything together
This is where we start to generate meaningful analyses with some probabilistic
modelling.  We keep the same three-state model, "alive", "progressive" and 
"dead", and begin again with a stationary example, before moving on to a more 
plausible (but complicated) non-stationary one.

### Example 4a: Probabilistic modelling, stationary model
First, we generate *sims* sets of PSA samples of the transition probabilities,
costs and health state utilities from appropriate distributions.  Here,
Dirichlet for probabilities, gamma for costs, and beta for health state
utilities.
```{r}
library(gtools)
sims <- 10
set.seed(34)
probs = cbind(rdirichlet(sims,c(80, 10, 10)),
              rdirichlet(sims,c(0, 7, 3)),
              rdirichlet(sims,c(0, 0, 1)))
colnames(probs) <- c("S1>S1", "S1>S2", "S1>S3",
                     "S2>S1", "S2>S2", "S2>S3",
                     "S3>S1", "S3>S2", "S3>S3")
stateCosts <- cbind(rgamma(sims, 10, 1/50),rgamma(sims, 1, 1/1000), 
                    rep(0, sims))
stateUtilities <- cbind(rbeta(sims, 800, 200), rbeta(sims, 50, 50),
                        rep(0, sims))
```

We then specify the state names, starting states, cycles, and convert the flat 
'probs' matrix using converttomatrices() helper function.  The Markov traces 
are calculated with a call to psamarkov().

```{r results = "hide"}
states <- c("alive", "progressive", "dead")
startingStates <- c(1, 0, 0)
cycles <- 10
transitionsMatrix <- converttomatrices(states, probs, cycles)
output <- psamarkov(transitionsMatrix, startingStates, cycles, states)
```

psamarkov() returns a list containing the means, upper and lower 95% credibility
limits for each state at each time point, as well as an array containing the 
entire raw data from each sim (the first two slices of which are shown below):

```{r}
print(output$means)
print(output$LCL)
print(output$UCL)

print(output$raw[,,1:2])
```

Now calculate the costs and outcomes using the sampled values of each (we'll use
the default transition period of 1 and discount rate of 3.5%):
```{r results = "hide"}
costs <- psavalues(output, stateCosts)
QALYs <- psavalues(output, stateUtilities, QALYs = TRUE)
```

And view some of the output:
```{r}
print(costs$undisc$mean)

print(costs$disc$total)
print(QALYs$disc$total)
```

We can retrieve a vector of estimated costs and QALYs in a format suitable for
use in [SAVI](https://github.com/Sheffield-Accelerated-VoI/SAVI) to calculate
value of information statistics:

```{r}
print(costs$disc$SAVI)
print(QALYs$disc$SAVI)
```

And assemble all the input parameters too (whilst making the column headers a 
little clearer - 'C' stands for cost and 'U' for utility):
```{r}
paramsSAVI <- cbind(probs,stateCosts,stateUtilities)
colnames(paramsSAVI) <- c(colnames(paramsSAVI)[1:9],
                          "C_S1", "C_S2", "C_S3", 
                          "U_S1", "U_S2", "U_S3")
head(paramsSAVI)
```

It's worth noting that the probability of moving from state 1 to states 1, 2 or
3 ("S1>S1", "S1>S2", and "S1>S3") are generated from the multinomial Dirichlet
distribution, and hence are correlated.  The EVPPI for the three parameters 
individually is therefore unhelpful, and these should be grouped together in
SAVI to calculate the EVPPI of "moving from state 1 to any other state".  The
same is true for the probabilities of moving from state 2 (and indeed state 3,
although this is the absorbing state and there is zero uncertainty).

### Example 4b: Probabilistic modelling, non-stationary model
The idea here is to use your favourite data-generating process to sample the
time-varying transition probabilities.  These are compiled into a wide matrix
called probs, where each row represents one complete set of simulated values.  

The order of the columns of probs is cycle number, state 1 to every other state,
state 2 to every other state and so on.  For example for a 2-state model with 3 
cycles, the complete order of columns would be:
C1_S1>S1, C1_S1>S2, C1_S2>S1, C1_S2>S2
C2_S1>S1, C2_S1>S2, C2_S2>S1, C2_S2>S2.  Where C = cycle, S = state, so C1_S1>S1
is the probability of transitioning from state 1 to state 1 in cycle 1.

Note cycles are numbered from 0, where the Markov chain is in its starting
state, so you only need to specify cycles-1 sets of transitions.

Suppose we're modelling the natural history of "Disease X" with the following
assumptions:

1. There are three health states, 'alive', 'progressive' and 'dead'.
1. The typical patient is a 60 year old female.
1. The model will run for 10 cycles.
1. Patients in the 'alive' state are at no increased risk of death compared with
the general population.
1. The probability of a patient developing progressive disease increases with
age, following a Weibull survival function.
1. Patients in the 'progressive' state have approximately 30% mortality risk
per annum (this is time-invariant to keep the example simple).

Set up a few model parameters reflecting (1)-(3) first:
```{r}
states <- c("alive", "progressive", "dead")
startingStates <- c(1, 0, 0)
startAge <- 60
cycles <- 10
```

Data source for (4) is lifetable data from [ONS](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/lifeexpectancies/datasets/nationallifetablesunitedkingdomreferencetables), 2016-18, females age 60-68.  
Given these are population level data, the standard error is small enough to 
enter these probabilities as constants in the model.
```{r}
lifeTable <- c(0.003548, 0.000224, 0.000127, 0.000098, 0.000073, 
               0.000081, 0.000075, 0.000060, 0.000060, 0.000062, 
               0.000059, 0.000076, 0.000069, 0.000078, 0.000101, 
               0.000119, 0.000153, 0.000152, 0.000218, 0.000196, 
               0.000197, 0.000224, 0.000219, 0.000220, 0.000226, 
               0.000260, 0.000252, 0.000286, 0.000330, 0.000314, 
               0.000374, 0.000394, 0.000482, 0.000500, 0.000545, 
               0.000586, 0.000654, 0.000738, 0.000720, 0.000846, 
               0.000882, 0.000993, 0.001051, 0.001183, 0.001328, 
               0.001436, 0.001540, 0.001700, 0.001823, 0.001935, 
               0.002136, 0.002363, 0.002581, 0.002756, 0.002952, 
               0.003265, 0.003621, 0.003896, 0.004324, 0.004730, 
               0.005104, 0.005600, 0.006303, 0.006832, 0.007346, 
               0.007995, 0.008860, 0.009523, 0.010379, 0.011361, 
               0.012601, 0.013781, 0.015916, 0.017545, 0.019298, 
               0.022011, 0.025052, 0.027787, 0.031365, 0.034408, 
               0.038922, 0.043938, 0.049786, 0.057500, 0.065049, 
               0.073792, 0.084250, 0.095303, 0.108358, 0.121616, 
               0.136979, 0.153256, 0.169425, 0.187195, 0.206281, 
               0.230368, 0.249110, 0.270828, 0.290344, 0.316440, 0.339686)
```

For (5) assume we have some long term observational data on a cohort of patients
with Disease X.  The Weibull function was shown to be the best fitting model and
has two parameters, shape and scale.  The mean (standard error of log of mean)
of each are 2 (ln1.05) and 0.001 (ln1.1) respectively.  The parameters are 
estimated from fitting the curve to the Kaplan-Meier plot of a cohort observed 
over time moving from 'alive' to 'progressive' disease.  The parameters are 
distributed log-normal.  We sample from these *sims* times:

```{r}
sims <- 10
set.seed(100)
WeibullSamples<- data.frame(shapes = rlnorm(sims, log(2), log(1.05)),
                      scales = rlnorm(sims, log(0.001), log(1.1)))
```

We then calculate *sims* Weibull functions with each sampled shape and scale
parameter, and convert the hazard to transition probabilities for ages 60-69.
(The transition probability between t(x-1) and t(x) is 1-exp(scale*((t(x-1)^shape 
- t(x)^shape))).

We store this in the matrix 'p_AtoP' ("probability of moving from alive to 
progressive").
```{r}
p_AtoP <- t(apply(WeibullSamples, 1,
                  function(y) sapply((startAge + 1):(startAge + cycles - 1),
                                     function(x) {
                                       1 - exp(y[2] * ((x - 1)^y[1] - x^y[1]))
                                     }
                                    )
                  )
            )
colnames(p_AtoP) <- paste0("p_AtoP_Age", startAge:(startAge + cycles - 2))
```

Next extract relevant lifetable data, and expand out replicating for each *sim*.
Save this in thematrix 'p_AtoD' ("probability of moving from alive to dead").
```{r}
lifetableExtract <- lifeTable[(startAge+1):(startAge + cycles - 1)]
p_AtoD <- matrix(rep(lifetableExtract, sims), nrow = sims, byrow = TRUE)
colnames(p_AtoD) <- paste0("p_AtoD_Age", startAge:(startAge + cycles - 2))
```

Each cycle, the probability of remaining in the 'alive' state is 1-sum(
probability of moving to 'progressive', probability of moving to 'dead'):
```{r}
StayingAlive <- 1 - p_AtoP - p_AtoD
colnames(StayingAlive) <- paste0("p_AtoA_Age", startAge:(startAge + cycles - 2))
```
We now have the transitions from the 'alive' state to 'alive', 'progressive' and
'dead' by transition period and sim.  Note as these are estimated separately, 
where the probabilities of death or progression are 'high', the probability of
remaining 'alive' may be negative.  Fortunately this doesn't occur in this 
example, but should this happen, the probabilities will need to be adjusted or
(ideally) jointly estimated.

Now set up the 'probs' matrix and populate with the transitions calculated so
far:
```{r}
nStates <- length(states)
probs <- matrix(data = NA, nrow = sims, ncol = nStates^2 * (cycles - 1))
colnames(probs) <- unlist(lapply(1:(cycles - 1),
                                 function(i) c(paste0("A>>A", i),
                                               paste0("A>>P", i),
                                               paste0("A>>D", i),
                                               paste0("P>>A", i),
                                               paste0("P>>P", i),
                                               paste0("P>>D", i),
                                               paste0("D>>A", i),
                                               paste0("D>>P", i),
                                               paste0("D>>D", i))
                                 )
                          )
probs[, seq(1, (cycles-1) * nStates^2, nStates^2)] <- StayingAlive
probs[, seq(2, (cycles-1) * nStates^2, nStates^2)] <- p_AtoP
probs[, seq(3, (cycles-1) * nStates^2, nStates^2)] <- p_AtoD
```
The transitions from progressive disease follow a simple Dirichlet(0,7,3), 
implying a mean 30% probability of death as per assumption (6), and we have no 
resurrection of the dead:
```{r}
library(gtools)

FromProgressive <- rdirichlet(sims, c(0, 7, 3))
for (i in 1:3) {
  probs[, seq(i + 3, (cycles - 1) * nStates^2, nStates^2)] <- FromProgressive[, i]
}

FromDead <- matrix(data = rep(c(0, 0, 1), sims), nrow = sims, byrow = TRUE)
for (i in 1:3) {
  probs[, seq(i + 6, (cycles - 1) * nStates^2, nStates^2)] <- FromDead[, i]
}
```
Now probs is set up, the model is calculated exactly the same as before:
reformatting probs and calculating the Markov traces, followed by eyeballing
some of the output:

``` {r results = "hide"}
transitionsMatrix <- converttomatrices(states, probs)
output <- psamarkov(transitionsMatrix, startingStates, stateNames = states)
```
```{r}
print(output$means)
print(output$LCL)
print(output$UCL)
```

Finally costs and health state utilities are sampled and applied to the Markov
trace:
```{r results = "hide"}
stateCosts <- cbind(rgamma(sims, 10, 1/50), rgamma(sims, 1, 1/1000), rep(0, sims))
stateUtilities <- cbind(rbeta(sims, 800, 200), rbeta(sims, 50, 50), rep(0, sims))

costs <- psavalues(output, stateCosts)
QALYs <- psavalues(output, stateUtilities, QALYs = TRUE)
```
```{r}
print(costs$undisc$mean)

print(costs$disc$total)
print(QALYs$disc$total)
```

As before, a vector of costs and QALYs can be extracted in [SAVI](https://github.com/Sheffield-Accelerated-VoI/SAVI) format:
```{r}
print(costs$disc$SAVI)
print(QALYs$disc$SAVI)
```

Assembling the input parameters in SAVI format is a little different from 
Example 4a.  For the Weibull survival function, Whilst we could simply add a 
column for everytransition at every time period, this is less useful and very 
inefficient.  What we really want are the sampled shape and scale parameters.  
There is no uncertainty in the life table data hence there is no point in 
including them. The sampled Dirichlet probabilities for transitions from the 
progressive state do need including as in Example 4a, and the same caveat is 
added: the EVPPI of the group of three should be examined, rather than for each
probability individually.

Finally we add the sampled costs and health state utilities too:
```{r}
paramsSAVI <- cbind(WeibullSamples, FromProgressive
                    ,stateCosts,stateUtilities)
colnames(paramsSAVI) <- c(colnames(paramsSAVI)[1:2],
                          "P_PtoA", "P_PtoP", "P_PtoD",
                          "C_S1", "C_S2", "C_S3", 
                          "U_S1", "U_S2", "U_S3")
head(paramsSAVI)
```


# Multi-core / parallel processing
Currently, using multiple cores increases time taken to compute by a factor of
around 2.  I think this is due to overhead from allocating *sims* to each core
sequentially one by one, whilst each individual calculation has a very quick 
processing time.  This may be solved by sending batches of sims to each core 
in turn rather than drip-feeding each core sim by sim. I will work on this in 
the future.  The non-functional parallel code is left in psamarkov() for the 
time being. 
