---
title: "validation-and-speed-check"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{validation-and-speed-check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Introduction
This vignette compares the results of the examples shown in the userguide with
a long-hand version created in MS Excel.  All excel files are available at ...
Note the Excel files are saved as macro-enabled files (*.xlsm) as the 
probabilistic analyses contain VBA code.  Deterministic analyses do not contain
code but are still saved as .xlsm for consistency.

For probabilistic analyses, a set of 10,000 sampled inputs is created in R with 
the stated number seed.  These are copied and pasted into excel, and VBA code 
used to calculate model results.  These are compared with the results generated
by R.  The computation time for each is reported.  Computation times in R with 
larger numbers of simulations are also reported.

# Computer spec
Analyses were conducted on an Intel(R) Core(TM) i7-6700K CPU @ 4.00GHz with 8
cores, running Windows 10 x64 with 24GB RAM:
```{r setup}
devtools::session_info()
```

# Test of Example code
## Example 1a
See file Example1a.xlsm
Difference in Markov traces = 0.  
Difference in total discounted QALYs = -2.1E-16.
Difference in total discounted costs = 0

Trivial differences in estimation of discounted QALYs and costs in some 
transition periods and total QALYs (at 16th and 13th significant figure 
respectively).  Most likely due to rounding errors.

## Example 1b
See file Example1b.xlsm
Difference in Markov traces = 0.  
Difference in total discounted QALYs = 0
Difference in total discounted costs = 0

Minor differences in estimation of discounted QALYs and costs (at 16th and 13th 
significant figure respectively).  Most likely due to rounding errors.

## Example 2a
See file Example2a.xlsm
Difference in Markov traces = 0.  
Difference in total discounted QALYs = 0
Difference in total discounted costs = 0

Minor difference in estimation of discounted cost in cycle 3 (at 13th 
significant figure).  Most likely due to rounding error.

## Example 2b
See file Example2b.xlsm
Difference in Markov traces = 0.  
Difference in total discounted QALYs = 0
Difference in total discounted costs = 0

Minor difference in estimation of discounted cost in cycle 3 (at 13th 
significant figure).  Most likely due to rounding error.

## Example 3
(these examples illustrate use of converttomatrices() helper function)

##Example 4a
See file Example4a.xlsm
three state, 10 cycle stationary model.
Maximum difference in mean discounted QALYs = <0.0000001
Maximum difference in mean discounted costs = <0.0000001
SD of difference in mean discounted QALYs = 2.4E-15
SD of difference in mean discounted costs = 3.3E-12

Computation time, 10,000 iterations:
MS Excel: 3 mins 27 seconds.
R (single core): 3.46 seconds.
R (8 cores):     


100,000 iterations (single core):              35.05 seconds.
100,000 iterations, 20 cycles (single core):   53.50 seconds.
1,000,000 iterations, 10 cycles (single core): 5 mins 9 seconds.
1,000,000 iterations, 20 cycles (single core): 8 mins 48 seconds.
100,000 iterations (8 cores):                  
100,000 iterations, 20 cycles (8 cores):       
1,000,000 iterations, 10 cycles (8 cores):     
1,000,000 iterations, 20 cycles (8 cores):     

##Example 4b
See file Example4a.xlsm
three state, 10 cycle non-stationary model.
Maximum difference in mean discounted QALYs = 
Maximum difference in mean discounted costs = 
SD of difference in mean discounted QALYs = 
SD of difference in mean discounted costs = 

Computation time, 10,000 iterations:
MS Excel: 
R (single core): 3.81 seconds

100,000 iterations (single core):              34.30 seconds
100,000 iterations, 20 cycles (single core):   
1,000,000 iterations, 10 cycles (single core): 6.088 mins
1,000,000 iterations, 20 cycles (single core): 
100,000 iterations (8 cores):                  
100,000 iterations, 20 cycles (8 cores):       
1,000,000 iterations, 10 cycles (8 cores):     
1,000,000 iterations, 20 cycles (8 cores):     
